Fast Sorting on Modern Computers

### 2005-07-27 docs/tech_report/heap.tex
reread the merge - updated results and created expected vs actual added a label to heap added a label to smallplot in report change multi mergesort to multi-mergesort
finished the comparison to lamarcs results for heapsort

A heap is a type of binary tree with several important properties. Firstly, the
the data stored by the parent must always be greater than the data stored by its
children. Secondly, in a binary tree there is no requirement that the tree be full.
In a heap, it is important that every node's left child exists before its right
child. With this property, a heap can be stored in an array without any wasted
space, and without any links to other nodes. A heap done in the fashion is
called an \n{implicit heap}.

To take advantage of the heap structure, it must be possible to easily traverse
the heap, going between parent and child, without the presence of links. This
can be done due to the nature of the indices of parents and their children. In
the case of an array indexed between $0$ and $N-1$, then a 
parent whose index is $i$ has two children with indices of
$2i+1$ and $2i+2$. Conversely, the index of a parent whose child has an index
of $j$ is $\lfloor{}(j+1)/2-1\rfloor{}$. It is also possible to index the array between $1$
and $N$, in which case the children have indices of $2i$ and $2i+1$, while the parent
has an index of $\lfloor{}j/2\rfloor{}$. Using this scheme can reduces instruction cost, but
results is slightly more difficult pointer arithmetic, which cannot be performed
in some languages\footnote{In C this can be done by \cc{Item a[] =
\&array[-1];}}.

@@

\section{Cache Heapsort}

LaMarca notes that the Repeated-Adds algorithm has a much lower number of cache
misses than Floyd's. The reason for this is simple. Consider the number of keys 
touched when a new key is added to the heap. In the case of Floyd's method, the
key being heapified is not related to the last key that was heapified. Reuse
only occurs upon reaching a much higher level of the heap. For large heaps, these
keys may be swapped out of the cache by this time. As a result, Floyd's method
has poor temporal locality.

@@

The height of a 2-heap is $\lceil{}log_2(N)\rceil{}$, where \textit{N} is the
number of keys in the heap. By changing to a 4-heap, the height is reduced to
$\lceil{}log_4(N)\rceil{} = \lceil{}\tfrac{1}{2}log_2(N)\rceil{}$. The height of
an 8-heap is half again the height of a 4-heap. However, the number of
comparisons needed in each level is doubled, since all the children will be
unsorted. LaMarca claims that as long as the size of the fanout is less than
12, a reduced instruction count will be observed.

@@

%TODO this isnt true is it? placing the nsentinel is cheaper than a bounds check
%(a write to a variable thats going to be written to in a second vs a comparison?)
%and there arent an extra logN instructions, theres logN times as many

@@

@@

As expected, the 4-heap performs better than the base heapsort. However, the
increased number of comparisons of the 8-heap increases the instruction count to
even higher than base heapsort, despite the reduction in the height of the heap.
This is shown in Figure \ref{heap instructions}.

The level 1 cache results, shown in Figure \ref{heap level 1 misses}, show the
difference between the two types of heap. The slope of the base heapsort result
is much greater than all the others, with the 8-heap having the fewest misses.
This continues into the level 2 cache, where the flattest slope belongs to the
8-heap. The difference between fully associative and direct mapped is only
minor, with the fully associative performing slightly better than the direct
mapped cache.

@@

The difference between the 4-heap and 8-heap is minor , and the difference for
larger fanout would be less still. The difference between the 2-heap and 4-heap
result is interesting though, the 2-heap has more branches, but less misses per
key.

>>>>

A heap is a type of binary tree with several important properties. Firstly, the
the data stored by the parent must always be greater than the data stored by its
children. Secondly, in a binary tree there is no requirement that the tree be full.
In a heap, it is important that every node's left child exists before its right
child. With this property, a heap can be stored in an array without any wasted
space, and the links are implied by a nodes position, rather than by a pointer
or reference. A heap done in the fashion is called an \n{implicit heap}.

To take advantage of the heap structure, it must be possible to easily traverse
the heap, going between parent and child, without the presence of links. This
can be done due to the nature of the indices of parents and their children. In
the case of an array indexed between $0$ and $N-1$, a parent whose index is
$i$ has two children with indices of $2i+1$ and $2i+2$. Conversely, the index of
a parent whose child has an index of $j$ is $\lfloor{}(j+1)/2-1\rfloor{}$. It is
also possible to index the array between $1$ and $N$, in which case the children
have indices of $2i$ and $2i+1$, while the parent has an index of
$\lfloor{}j/2\rfloor{}$. Using this scheme can reduces instruction cost, but
results is slightly more difficult pointer arithmetic, which cannot be performed
in some languages\footnote{In C this can be done by \cc{Item a[] =
\&array[-1];}}.

@@

\section{Cache Heapsort}
\label{cache heapsort}
LaMarca notes that the Repeated-Adds algorithm has a much lower number of cache
misses than Floyd's. The reason for this is simple. Consider the number of keys 
touched when a new key is added to the heap. In the case of Floyd's method, the
key being heapified is not related to the last key that was heapified. Reuse
only occurs upon reaching a much higher level of the heap. For large heaps, these
keys may be swapped out of the cache by this time. As a result, Floyd's method
has poor temporal locality.

@@

The height of a 2-heap is $\lceil{}log_2(N)\rceil{}$, where \textit{N} is the
number of keys in the heap. By changing to a 4-heap, the height is reduced to
$\lceil{}log_4(N)\rceil{} = \lceil{}\tfrac{1}{2}log_2(N)\rceil{}$. The height of
an 8-heap is reduced again to $\lceil{}log_8(N)\rceil{} =
\lceil{}\tfrac{1}{3}log_2(N)\rceil{}$. However, the number of comparisons needed
in each level is doubled, since all the children will be unsorted. LaMarca
claims that as long as the size of the fanout is less than 12, a reduced
instruction count will be observed.

@@

@@

LaMarca reports base heapsort has very poor temporal locality, and that the
cache heapsort will have half the cache misses of base heapsort.

@@

As expected, the 4-heap performs better than the base heapsort. However, the
increased number of comparisons of the 8-heap increases the instruction count to
even higher than base heapsort, due to the disparity between the increase in
comparisons in set of children, and the lesser reduction in the height of the
heap. This is shown in Figure \ref{heap instructions}.

The level 1 cache results, shown in Figure \ref{heap level 1 misses}, show the
difference between the two types of heap. The slope of the base heapsort result
is much greater than all the others, showing base heapsort's poor temporal
locality, with the 8-heap having the fewest misses. This continues into the
level 2 cache, where the flattest slope belongs to the 8-heap. The difference
between fully associative and direct mapped is only minor, with the fully
associative performing slightly better than the direct mapped cache.

@@

The difference between the 4-heap and 8-heap is minor, and the difference for
larger fanout would be less still. The difference between the 2-heap and 4-heap
result is interesting though, the 2-heap has more branches, but less misses per
key.
